---
output:
  
  html_document: default
  word_document: default
---
ANIKO LENGYEL - WHITE WINE DATA SET ANALYSIS
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(GGally)
library(doBy)
library(corrplot)

```


I load the data and get the first seven rows.

```{r echo=FALSE, Load_the_Data}

wine_data <- read.csv('wineQualityWhites.csv')

head(wine_data, 7)

```

Looking at the data structure and data types.

```{r echo=FALSE}

str(wine_data)

```

Getting the summary of the dataset to get information about the missing rows 
(if there are any).

```{r echo=FALSE}


sum(is.na(wine_data))

```

###Introduction:
I found this public dataset among the Udacity's data set options.
It catched my eyes because I am interested in wine making and wanted 
to know more about its background.
The dataset consists of 13 variables and 4898 observations, 
integer and numeric data types, containing data about different features of 
white wines, for example acidity, sugar and alcohol contain and quality.
I plan to analyze the relation between the different components and wine 
quality.

# Univariate Plots Section

At first, I create a bar plot to get a quick view about the distribution 
of the quality rates and I run a summary to generate some summary statistics.

```{r echo=FALSE}

ggplot(wine_data, aes(factor(quality))) +
  geom_bar()

```

```{r echo=FALSE}

summary(wine_data$quality)

```

I got a normally distributed histogram, most wine have a quality between 5 
and 7, the tallest clusters of bars is 6, representing the most common quality.
Based on my summary, the the mean value is 5.878, the min quality is 3 
and the max quality is 9.

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= pH, data = wine_data)

```

```{r echo=FALSE}

summary(wine_data$pH)

```

The histgoram of pH is also normally distibuted and concentrated around 3.15. 
The min and max values are 2.72 and 3.82, the median is 3.18.

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= sulphates, data = wine_data)

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= sulphates, data = wine_data, xlim = c(0.35, 0.9))

```

```{r echo=FALSE}

summary(wine_data$sulphates)

```

The histogram of suplhate content is right skewed - the most oftten occuring 
values are between 0.4 and 0.5, the peak is about 400. I used an xlim function 
to improve readability and remove outstanding values.

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= density, data = wine_data, xlim = c(0.987, 1.002))

```

```{r echo=FALSE}

summary(wine_data$density)

```

The the density hitogram looks normal distributed at the first sight, but it is 
a little skewed to the left. The difference between min and max values are less
than 0.5. 

On the second plot, I used the parameter xlim to handle the outliers 
and make the plot more readable.

```{r echo=FALSE}

qplot(x= fixed.acidity, data = wine_data, binwidth = 0.2)

``` 

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= fixed.acidity, data = wine_data, binwidth = 0.2, xlim = c(4, 10))

``` 

```{r echo=FALSE}

summary(wine_data$fixed.acidity)

```

The ditribution of fixed acidity has its peak around 7 and skeewed 
to the right. I removed the outliers with xlim on the second plot.

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= volatile.acidity, data = wine_data, binwidth = 0.02)

``` 

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= volatile.acidity, data = wine_data, binwidth = 0.02, xlim= c(0.08, 0.7))

``` 

```{r echo=FALSE}

summary(wine_data$volatile.acidity)

```

The histogram of volatile acidity is concentrated around 0.2 and skewed to the 
right. The most values are lacated between 0.15 and 0.325. I transformed the 
second plot with xlim to limit the outliers.

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= residual.sugar, data = wine_data)

``` 

```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x= residual.sugar, data = wine_data) +
  scale_x_log10()

``` 

```{r echo=FALSE}

summary(wine_data$residual.sugar)

```

The distribution of residual sugar is skewed to the right. Based on the summary
and the histogram, the distribution of the sugar content is relatively divided: 
the min value is 0.6 and the max value is 65 - ten times the min value! The 
values first three quarter are less than 9.9, indicating that there is only 
a few wines which contains less sugar. This also corresponds to the histogram:
I can hardly see any value above 20. 
There were some outliers on the first residual sugar plot, so I decided to use 
scale_x_log_10() to handle them. 

```{r echo=FALSE}

qplot(x= alcohol, data = wine_data, binwidth = 0.5)

``` 


```{r echo=FALSE}

summary(wine_data$alcohol)

```

The distribution of alcohol is right skewed, most wine have their alcohol content
between 8.5 and 11. There is no wine with lower alcohol content that 8,
and the maximum alcohol content is a little above 14. I setted the bidwidt to
0.5 for a better look.

# Univariate Analysis

### What is the structure of your dataset?

The dataset consists of 13 variables and 4898 observations 
and contains data about different features of white wines, including rates 
of acidity, pH value, sugar contain and quality rates. 
The variables are all quantitative values, represented by integer and numeric 
data types. According to my calculation, there are no any missing values.

### What is/are the main feature(s) of interest in your dataset?

I would like to determine which features are the most important in predicting 
the wine quality, so I plan to examine the possible correlaton 
between wine quality and other components, like residual sugar, alcohol content, 
acidity and density.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

As far I know, sugar content can reasonably influence the quality of the wine.
The high quality wines have lower sugar content while the lower quality often
comes with a higher sugar content. I will also analyze the correlation
between alcohol content, density and quality.

### Did you create any new variables from existing variables in the dataset?

I did not create any new variables in the dataset. However, I was wondering 
about creating three bins based on wine quality (like bad-average-good) but 
I decided to  make a more detailed analysis on quality.

### Of the features you investigated, were there any unusual distributions? \

The residual sugar and volatile acidity data have some unusual distributions - 
both of the histograms are strongly right skewed. I wonder if these components
have something to do with wine quality.

Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Since I got a tidy dataset, I did not need to perform any cleaning 
operation or other kind of adjusting.


# Bivariate Plots Section

I plotted the variables of the dataset to get a quick
visualization about which values are relevant in predictiong wine quality.
Based on the matrix,  presume alcohol content to correlate with quality. 
Moreover, I also found some other interesing connections worth looking at it. I
sat up a 0.95 confidence level and a 0.05 significance and marked the 
corresponding variables with a black X.

```{r echo=FALSE, fig.width=6, fig.height=6}

corr_wine <- cor(wine_data)

sign <- cor.mtest(wine_data, conf.level = .95)

corrplot(corr_wine, method = "circle", cl.pos = "b",
         p.mat = sign$p, sig.level = .05, tl.col = "black")

```

I made a scatter plot diagram about the relationship between fixed acidity and 
pH. According to the visualization, the lower the fixed acidity, the higher the 
pH value, what absolutely makes sense. I also counted the correlation between 
the two variables and found a moderate negative relationship (-0.426) and used
stat_smooth() function to represent it.

An article about pH and wine acidity which I used to confirm and better 
understand my findings: 
http://winefolly.com/review/understanding-acidity-in-wine/

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = pH, y = fixed.acidity), data = wine_data)+
  geom_point(alpha = 1/3)+
  scale_y_continuous(limits = c(3, 12)) +
  stat_smooth()

```



```{r echo=FALSE}

cor(wine_data$pH, wine_data$fixed.acidity)

```

I created a scatter plot about sulphates and alcohol content. I did not found any 
correlation between the two vales, according to my correlation coefficient 
number, their correlation is just -0.017.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = sulphates, y = alcohol), data = wine_data)+
  geom_point(alpha = 1/3)+
  stat_smooth()

```

```{r echo=FALSE}

cor(wine_data$sulphates, wine_data$alcohol)

```

I wondered if residual sugar can influence the density of a wine, so, I plotted 
the correlation between density and residual sugar and counted the 
correlation coefficient. I found a strong positive correlation between the two
variables (0.839) - as sugar content increases, the density increases. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = residual.sugar, y = density), data = wine_data)+
  geom_point(alpha = 1/5) +
  scale_x_continuous(limits = c(0, 30)) +
  scale_y_continuous(limits = c(0.9875, 1.005)) +
  stat_smooth()

```

```{r echo=FALSE}

cor(wine_data$density, wine_data$residual.sugar)

```

A created a scatter plot about the alcohol content and density and counted the 
correlation coefficient. According to my results, there is a strong negative
correlation between the two component (-0.7801) - the higher the alcohol content
is, the lower the density.


```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = alcohol, y = density), data = wine_data)+
  geom_point(alpha = 0.2) +
  scale_y_continuous(limits = c(0.985, 1.005)) +
  geom_smooth()

```


```{r echo=FALSE}

cor(wine_data$alcohol, wine_data$density)

```

I created a jittering scatter plot about the correlation between residual sugar 
and quality. I did not find any correaltion between the two and the correlation 
coefficient was also quite week, only -0.0976. According to my results, I can 
state that there is no evidence that residual sugar would influence quality. 
This outcome does not match up with my expectations, I thought
the best and most prestigous wines are dry and have a lower sugar content on 
the average.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = quality, y = residual.sugar), data = wine_data)+
  geom_jitter(alpha = 0.2) +
  scale_y_continuous(limit = c(0, 25))

```

```{r echo=FALSE}

cor(wine_data$quality, wine_data$residual.sugar)

```

I used the summaryBy function from doBy package to summarize the residual sugar 
content by quality as I was curious about the more precise mean values.

Source: https://www.rdocumentation.org/packages/doBy/versions/4.5-15/topics/summaryBy

```{r echo=FALSE}

summaryBy(residual.sugar~quality, data=wine_data, FUN=c(mean))

```

I used boxplots to visualize the correlation between quality and alcohol content.
I counted the correlation coefficient and got a positive moderate correlation 
(0.435). The boxplots also show that higher quality wines have higher median 
alcohol content (above 11 and even 12), while lower quality wines have their 
median alcohol content around 10. The minimum and maximum values also follow 
this pattern: the worse wines are their maximum around 12-13 while the betters 
around 14.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = factor(quality), y = alcohol), data = wine_data)+
  geom_boxplot() +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

```

```{r echo=FALSE}

summaryBy(alcohol~quality, data=wine_data, FUN=c(mean, min, max))

```

```{r echo=FALSE}

cor(wine_data$alcohol, wine_data$quality)

```

```{r echo=FALSE}

summaryBy(alcohol~quality, data=wine_data, FUN=c(mean))

```

I read in the dataset description that too high volatile acidity can lead to
an unpleasent, vinegar taste and ruin qualityso I made boxplots about the 
correlation between volatile acidity and quality.The calculated correlation
between volatile acidity and quality does not match up with my expectation. 
However, if we look at the boxplots, and compare 
the mins, maxes and look at the values between the first and third quantile, 
we can find out that lower quality wines still have a higher volatile acidity 
content.
I also used a jittering plots to visualize the distribution of the values
and marked the means with a red dot.

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = factor(quality), y = volatile.acidity), data = wine_data) +
  geom_jitter(alpha = 0.2) +
  scale_y_continuous(limits = c(0.08, 0.7)) +
  geom_boxplot(alpha = 0.3, color = 'blue', lwd = 1.05) +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 16, 
               size = 3)

```

```{r echo=FALSE}

cor(wine_data$volatile.acidity, wine_data$quality)

```

```{r echo=FALSE}

summaryBy(volatile.acidity~quality, data=wine_data, FUN=c(mean, min, max))

```

I plotted the correlation between wine quality and density and found a negative 
relationship on the edge of weak and moderate categories (-0.3071). However, 
according to my groupped calculation, the best wines have the lowest mean 
density.

Source for interpreting correlation coefficient:
http://www.dummies.com/education/math/statistics/how-to-interpret-a-correlation-coefficient-r/

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = quality, y = density), data = wine_data)+
  geom_jitter(alpha = 0.3) +
  scale_y_continuous(limits = c(0.987, 1.005))

```

```{r echo=FALSE}

cor(wine_data$quality, wine_data$density)

```


```{r echo=FALSE}

summaryBy(density~quality, data=wine_data, FUN=c(mean, min, max))

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

My main feature of the interest was the quality of the wine, I examined which 
components can influence the wine quality in the dataset and how strong is
the correlation between quality and other features. I found out that alcohol
content and density can play a lot more important role in quality.

- Residual sugar does not correlate with wine quality at all. The mean values 
vary between 4 and 7.5 and the best wines have the lowest mean sugar content 
(4.12). Hovewer, we can not predict quality based on only the residual sugar.

- The variation of mean alcohol content among the different qualities is about 
2%. The best wine have the highest mean alcohol content (12.18). The top
three wine have their mean alcohol above 11 while the worst above 11.

- The variation among mean density values is less than 0.01. There is negative
0.307 correlation between quality and density that I would rather interpret as 
negligible and weak.

- Volatile acidity has the strongest correlation with wine quality among 
other acidity indicators, the worst wines have a higher volatile acidity
proportion.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I observed a strong correlation between residual sugar and density, indicating 
that wines with more density have more residual sugar content. I also found 
interesting the strong negative correlation between density and alcohol 
content - the more alcohol the wine contains the lower the density is. As a
non-professional, I would expect quite the opposite.

### What was the strongest relationship you found?

The strongest relationship I found was between alcohol content and residual
sugar, it is 0.8389.

# Multivariate Plots Section

I created a scatterplot to investigate the interactions between residual sugar, 
density. I grouped and colored the dots by quality to show the correlation
between density and quality, too. As a result, we got a strong positive 
correlatedscatter plot where dots, representing the better wines can be 
found on the bottom of the plot. I also used regression lines and dark theme 
to make the trends easier to understand.

```{r echo=FALSE, Multivariate_Plots, warning=FALSE, message=FALSE}

ggplot(aes(x = residual.sugar, 
           y = density, color = factor(quality)), data = wine_data) +
      geom_point(alpha = 0.5, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq') +
  scale_y_continuous(limits = c(0.987, 1.0025)) +
  scale_x_continuous(limits = c(0, 20)) +
  theme_dark()

```

I make a plot about the possible trend between pH and sulphates, colored by 
quality. I used the stat_smooth() function to stress the interesting 
trend what can be even interpreted as nonlinear. The lower and higher suplhate
content correlates with relative high alcohol, but the average suplhate content
indicates lower alcohol proportion. If we look at the colors, we can also notice
the strong positive trend of distribution of quality. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(aes(x = sulphates, y = alcohol), data = wine_data)+
  geom_point(aes(color = factor(quality))) +
  stat_smooth() +
  scale_colour_brewer(palette = "Reds")

```

I created a box plot to visualize the trend between quality, alcohol and 
residual sugar. I made buckets to mark and separate the different levels of 
residual sugar proportions and get a better readable visualization.
We can observe the variation on alcohol content by quality: 
the better the wine the higher the median alcohol content is for each level
of quality (the medians are marked by a red star).

The opposite goes for the residual sugar, as wines with higher alcohol content 
contain less sugar. This tendency also reflects on one important part of wine 
making, as "residual sugar is the level of glucose and fructose (grape sugars) 
that are not converted into alcohol during fermentation".

Source: http://winefolly.com/tutorial/wines-from-dry-to-sweet-chart/


```{r echo=FALSE, warning=FALSE, message=FALSE}

wine_data$residual.sugar_bucket = cut(wine_data$residual.sugar,
                            c(0, 10, 20, 40, 65.800 ))

ggplot(aes(x = factor(quality), y = alcohol ), data = wine_data) + 
  geom_boxplot( aes(fill= residual.sugar_bucket))  +
  scale_fill_brewer(type='seq',
                   guide=guide_legend(title=wine_data$residual.sugar_bucket))+
  stat_summary(fun.y = "median", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Alcohol content, residual sugar and density on Plot 1 strengthened
each other as I started using color to hightlight some aspects of the data. This
is also true in case of Plot 3, where I compared alcohol content and residual
sugar to quality.

### Were there any interesting or surprising interactions between features?

I found really interesting the interaction between alcohol an residual sugar 
content and its reason. I never thought about this part of wine making before.
I also found an unusual, non linear-looking correlation
between alcohol and sulphates. As a non-professional, I made a little research 
to find out if there are any information about their relationship, but I have 
not found anything yet, it can also be a random exception. Suplhates are not 
involved in wine making naturally, but some wine maker use it to correct mineral 
deficiencies.

Source: 
http://www.foodsmatter.com/allergy_intolerance/sulphites/articles/sulphates_sulphites.html

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One}

plot1 <- ggplot(aes(x = factor(quality), y = volatile.acidity), data = wine_data)+
  geom_boxplot(aes(fill = "c13016"), show.legend = FALSE) +
  xlab('Quality (Score Between 0 and 10)') +
  ylab('Volatile Acidity (acetic acid - g/dm³)') +
  ggtitle('The Relationship Between Quality and Volatile Acidity') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 3)

plot1 +
  theme(
  panel.background = element_rect(fill = "#d2c489",
                                colour = "#d2c489"))

```

### Description I:

The visualisation shows the correlation between volatile acidity and quality. 
We can not necessarily observe a linear trend among the quality ranges but
the worst wines tend to have a higher volatile acidity, what 
absolutely makes sense: wines with higher volatile acidity could taste bad. If
we compare the values between the first and third quantile we can see that the
worst wines (quality 3 and 4) have higher volatile acidity while the better ones
(quality 5-7) have lower values. This trend does not corresponds to the best
wines (quality 8 and 9).


### Plot Two
```{r echo=FALSE, Plot_Two, warning=FALSE, message=FALSE, fig.width= 8.5, fig.height= 6}

plot2 <- ggplot(aes(x = factor(quality), y = alcohol ), data = wine_data) + 
  geom_boxplot( aes(fill= residual.sugar_bucket))  +
  scale_fill_brewer(type='seq', palette = "Reds",
                   guide=guide_legend(title=wine_data$residual.sugar_bucket))+
  xlab('Quality (Score Between 0 and 10)') +
  ylab('Alcohol (% by Volume)') +
  ggtitle('The Correlation Between Quality and Alcohol Content') +
  stat_summary(fun.y = "median", 
               geom = "point", 
               color = "black", 
               shape = 8, 
               size = 3)

plot2 +
  theme(
  panel.background = element_rect(fill = "#C0C0C0",
                                colour = "#C0C0C0")) +
  guides(fill=guide_legend(title="Residual Sugar 
(g/dm³)"))

```

### Description II:

I used the previously bucketed values and box plots to show the most important 
factor which influence wine quality - alcohol. We can also get a more detailed 
perspective about the correlation, thanks to the categorized correlation lines. 
The better wines have a higher alcohol content and lower residual sugar.
Previously we found that residual sugar does not correlate directly with
wine quality, however, sugar and alcohol have a strong correlation and alcohol
indicates wine quality very strongly.

### Plot Three

```{r echo=FALSE, Plot_Three, warning=FALSE, message=FALSE, fig.width= 8, fig.height=6}

plot3 <- ggplot(aes(x = residual.sugar, y = density, color = factor(quality)), 
                data = wine_data)+
  geom_point(alpha = 0.5) +
  scale_y_continuous(limits = c(0.987, 1.0025)) +
  scale_x_continuous(limits = c(0, 25)) +
  xlab('Residual Sugar Content (g/dm³)') +
  ylab('Density (g/cm³)') +
  scale_color_brewer(palette = "Reds") +
  geom_smooth(method = "lm", se = FALSE, size=1)


plot3$labels$colour <- "Quality 
(Score between 1 and 10)"

plot3 +
  theme(
  panel.background = element_rect(fill = "#C0C0C0",
                                colour = "#C0C0C0"))

```

### Description III:

Residual sugar and density are highly correlated. Quality -marked by color-
also tend to correlate with density. The lower the quality is, the higher the 
density.

------

# Reflection

My main feature of interest was wine quality and I could find some variables
which tend to influence this factor, like alcohol content and density. In the 
first part I made some simple hitograms about the main factors and other 
variables as well. In the part of bivariate analysis, I tried to get a better
knowledge on the factors which influence wine quality. I was suprised
that residual sugar does not correlate with wine quality. I also could make some
other interesting explorations which bring further questions and could be 
analyzed in the future, especially the relationship of sulphates and alcohol.
In the third part I highlighted some of my main findings, and reflected the 
main components that influence quality.

The project work took me more time than I previously estimated: one
of the reason is that I forgot about an important rule: the devil is in the 
details. It took time to choose the best corresponding plots and to detect the 
variables that influence quality. I had to examine each variable separatly 
while not forgetting the whole picture which I found really challenging and 
exciting. 